---
title: "Homework 02: Explore Gapminder and use dplyr"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting ups

Before doing anything exciting with `gapminder` and `dplyr`, first we should get the packages installed and loaded.

```{r results = 'hide'}
install.packages("gapminder")
install.packages("tidyverse")
library(gapminder)
suppressPackageStartupMessages(library(tidyverse))
```

## Smell test the data

- Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
```
It is a data.frame, but when using `typeof`, the output is always base R type.

- What is its class?
```{r}
class(gapminder)
```
As mentioned above, the class of `gapminder` is data.frame.

- How many variables/columns?
```{r}
ncol(gapminder)
```

- How many rows/observations?
```{r}
nrow(gapminder)
```

- Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?
```{r}
dim(gapminder)
str(gapminder)
```

- What data type is each variable?
```{r}
head(gapminder)
```
We can find the data types in the row below the variable names.

## Explore individual variables

- What are possible values (or range, whichever is appropriate) of each variable?
```{r}
summary(gapminder $ continent)
summary(gapminder $ pop)
```

- What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

`summary` function can show us many things mentioned above, or we can check the distribution of the data by bar plot.
```{r}
gapminder %>% 
  ggplot(aes(x = pop)) + geom_histogram(bins = 100) + scale_x_log10()
```

## Explore various plot types

- A scatterplot of two quantitative variables.
```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent, size = year)) + geom_point(alpha = 0.5) + scale_x_log10()
```

- A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
```{r}
gapminder %>% 
  ggplot(aes(x = lifeExp)) + geom_density()
```

- A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.
```{r}
gapminder %>% 
  ggplot(aes(x = continent, y = lifeExp, color = year, size = pop)) + geom_violin() + geom_jitter()
```

## Use `filter()`, `select()` and `%>%`

With `filter` function we can create a data subset and do some exploration.
```{r}
gapminder %>% 
  filter(continent == "Americas" & pop > 1e7) %>% 
  ggplot(aes(x = year, y = lifeExp, color = country)) + geom_point(aes(size = pop), alpha=0.6) + geom_line()
```

## But I want to do more!

- The correct form of boolean expression
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
The right way is using `%in%` expression here.
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

- Use `knitr::kable` to present numerical tables in a more attractive form.
```{r}
knitr::kable(filter(gapminder, continent %in% c("Asia", "Americas") & year == 2002 & pop > 3e7))
```

- More explorations about `dplyr`
```{r, results='hide'}
gapminder %>% 
  filter(continent == "Americas" & pop > 2e7) %>% 
  ggplot(aes(x = year, y = gdpPercap)) + geom_point(aes(color = country, size = lifeExp), alpha = 0.6) + geom_smooth()
```